---
title: "Problem Set 3"
author: "Arihaan Barua"
date: now
format: html
editor: source
self-contained: true
---

## Question 1

Part (a)
```{r}
#| echo: false
library(dplyr)
library(Lahman)
(bat_300_300 <- Batting %>%
  group_by(playerID) %>%
  summarise(
    total_HR = sum(HR, na.rm = TRUE),
    total_SB = sum(SB, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(total_HR >= 300, total_SB >= 300) %>%
  left_join(People %>% select(playerID, nameFirst, nameLast), by = "playerID") %>%
  select(nameFirst, nameLast, total_HR, total_SB) %>%
  arrange(desc(total_HR), desc(total_SB), nameLast, nameFirst))
```

Part (b)
```{r}
#| echo: false
(pitch_300_3000 <- Pitching %>%
  group_by(playerID) %>%
  summarise(
    total_W  = sum(W,  na.rm = TRUE),
    total_SO = sum(SO, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(total_W >= 300, total_SO >= 3000) %>%
  left_join(People %>% select(playerID, nameFirst, nameLast), by = "playerID") %>%
  mutate(player = paste(nameFirst, nameLast)) %>%
  select(player, total_W, total_SO) %>%
  arrange(desc(total_W), desc(total_SO), player))
```

Part (c)
```{r}
#| echo: false
(top5_hits <- Batting %>%
  group_by(playerID) %>%
  summarise(
    total_H = sum(H, na.rm = TRUE),
    start_year = min(yearID, na.rm = TRUE),
    end_year   = max(yearID, na.rm = TRUE),
    career_years = end_year - start_year + 1,
    .groups = "drop"
  ) %>%
  left_join(People %>% select(playerID, nameFirst, nameLast), by = "playerID") %>%
  arrange(desc(total_H)) %>%
  slice(1:5) %>%
  select(nameFirst, nameLast, career_years, total_H))
```

## Question 2

Part (a)
```{r}
#| echo: false
height_gpa <- read.table(
  "https://online.stat.psu.edu/onlinecourses/sites/stat501/files/data/heightgpa.txt",
  header = TRUE
)
plot(
  gpa ~ height,
  data = height_gpa,
  xlab = "Height (inches)",
  ylab = "GPA",
  main = "Scatterplot of GPA vs Height"
)
```
There appears to be a weak positive linear relationship, but the datapoints are pretty evenly spread out.

Part (b)
```{r}
(cor(height_gpa$height, height_gpa$gpa))
```

Part (c)
```{r}
model <- lm(gpa ~ height, data = height_gpa)
summary(model)
```
The estimated slope represents the change in mean GPA for every one-inch increase in height. In this dataset, the slope is very small, suggesting that height doesn't much effect on GPA.

Part (d)
```{r}
#| echo: false
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))
```

Part (e)
```{r}
summary(model)$r.squared
```
The R^2 value is very small, meaning that height explains only a small proportion of the variability in GPA. This means that almost all variation in GPA is due to other factors.

Part (f)

This data set shows that there is little to no meaningful relationship between a student’s height and their GPA. Both the correlation coefficient and R^2 value indicate that height is not a useful predictor of academic performance in this sample.